---
title: "Sticks Kebab Shop Segmentation"
author: "Team 1"
date: "4/12/2020"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readxl)
library(tidyr)
fname = "C:\\Users\\mbeck\\Google Drive\\MAR 653\\Homework\\M0866Xv1hw1assignmentfile.xlsx"
segmentation_data <- read_excel(fname,sheet="data for segmentation")
# Omit NULL Responses
seg_data_wip <-na.omit(segmentation_data)
colnames(seg_data_wip)[3:6] <-c("careful_planner","spender","local_oriented","health_oriented")

```

## How do people choose a fast food restaurant to visit?

- What is important, location, price, assortment, or cuisine?
- Who are Stick's customers, and what are their motivations for visting sticks?
- What does the survey data tell us about differences between customers and
Noncustomers?

- What survey questions would you use to identify the customer segments?

## How many customer segments can you estimate from the survey data?

- What are the profiles of the customer segments?
- Which customer segments should Sticks target?
- Bullet 3

## Recommendation for the location of the next Sticks Kebob Shop 

- based on the segmentation analysis and the demographic profiles of the locations provided.



```{r cars, echo = TRUE}
summary(cars)
```

## Determining the optimal number of clusters

```{r pressure}

# Determine optimal number of clusters
k.max <- 8
data <- seg_data_wip
wss <- sapply(1:k.max, 
              function(k){
                kmeans(data, k, nstart=500,iter.max = 8 )$tot.withinss
                })
wss
plot(1:k.max, wss,
     type="b", pch = 19, frame = FALSE, 
     xlab="Number of clusters K",
     ylab="Ratio of within Cluster to between Cluster Variance")


# Using optimal number of clusters, Run the analysis and add the cluster indicators.
cluster_segments <- kmeans(seg_data_wip[,3:6],3)
cluster_segments
seg_data_wip$cluster <- cluster_segments$cluster

write.csv(seg_data_wip, file = "C:\\Users\\mbeck\\Google Drive\\MAR 653\\Homework\\clustered_segments.csv")


```

